Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent

### 2023-06-27 manuscript/main.tex
rework introduction

@@

A pangenome graph layout is the arrangement of nodes and edges in n-dimensional space such that a human accessible representation of genetic variation across multiple genomes can be drawn.
The layout challenge is to find the best node positions in order to minimize overlapping nodes or edges, reduce edge crossings, and promote intuitive understanding by displaying structure and important features in a human readable way.
\FIXME{
	Existing layout algorithms range from simple, force-directed methods to more complex, hierarchical layouts, each presenting unique strengths and drawbacks.
	Force-directed methods offer an aesthetic layout based on a physics simulation but may struggle with scalability for larger graphs (XXX WE NEED CITATIONS XXX).
	In contrast, hierarchical layouts handle layered relationships and dependencies, yet can exhibit limitations when dealing with non-hierarchical data (XXX WE NEED CITATIONS XXX)
}

With the ever-growing scale and complexity of genomic data, there is a need for advanced layout algorithms that can optimally balance factors such as
efficiency, scalability, readability, and the ability to reveal significant biological patterns.
Pangenome graphs embed linear sequences as paths in the graph, but to our knowledge, no algorithm takes into account this biological information during the layout process.
We present a new pangenome graph layout algorithm which uses path-guided stochastic gradient descent %\citep{Zheng2019}
 (PG-SGD) to move pairs of nodes in parallel. ( \FIXME{WE NEED TO SAY SOMETHING ABOUT HOGWILD HERE?!}). \\
\FIXME{A LITTLE BIT MORE ABOUT THE SGD APPROACH?} \\
We exemplify how our implementation is a key step in general pangenome analyses such as pangenome graph buildiong and analysis (\FIXME{CITE PGGB, ACROCENTRIC PAPER, HPRC PAPER, ODGI PAPER}). \\
\FIXME{2D: waragraph + gfaestus for interactive visualization}.

@@

    \begin{appendices}
        \section{Quantization}
        \FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.}
    \end{appendices}

>>>>

A pangenome graph layout is the arrangement of nodes and edges in n-dimensional space such that a human accessible representation of genetic variation across multiple genomes can be drawn. 
\FIXME{We may have to talk about paths and nucleotides here. Or we introduce this later, not sure.}
The layout challenge is to find suitable node coodinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote intuitive understanding by displaying structure and important features in a human readable way. 
One popular graph layout approach is force-directed graph drawing which produces aesthetically-pleasing layouts. It is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations can circumvent that \citep{Zheng2019}. Typically, force-directed layouts are hard to compute. HOGWILD! \citep{Recht2011} SGD presents a parallelizable solution. In practice, multidimensional scaling (MDS) using pairwise nodes distances as a (dis)similarity measurement is applied to minimize the energy/stress/secs of a graph.
Since pangenome graphs embed linear sequences as paths in the graph, a reasonable distance metric would be the path nucleotide distance between a pair of nodes. 
But to our knowledge, no algorithm takes into account this biological information during the layout process.

We present a new pangenome graph layout algorithm which uses path-guided stochastic gradient descent %\citep{Zheng2019}
(PG-SGD) to move pairs of nodes in a HOGWILD! parallel manner. The algorithm aims to adjust the graph layout to best reflect the path nucleotide positions.
We exemplify how our implementation is able to produce human readable layouts. Furthermore, the 1D PG-SGD already is a key step in pangenome graph construciton \citep{Liao2023, Garrison2023} and analysis \citep{Guarracino2022, Guarracino2023}.

\FIXME{2D: waragraph + gfaestus for interactive visualization? Not sure how to add this here. I am afraid the readers might think we also implemented these tools. Better in the discussion?}. \\
% We may want to adjust some citations using https://tex.stackexchange.com/questions/585942/how-to-display-the-first-two-authors-in-a-citation-call-out.

%\FIXME{
%	Existing layout algorithms range from simple, force-directed methods to more complex, hierarchical layouts, each presenting unique strengths and drawbacks.
%	Force-directed methods offer an aesthetic layout based on a physics simulation but may struggle with scalability for larger graphs (XXX WE NEED CITATIONS XXX).
%	In contrast, hierarchical layouts handle layered relationships and dependencies, yet can exhibit limitations when dealing with non-hierarchical data (XXX WE NEED CITATIONS XXX)
%}

%With the ever-growing scale and complexity of genomic data, there is a need for advanced layout algorithms that can optimally balance factors such as
%efficiency, scalability, readability, and the ability to reveal significant biological patterns. \FIXME{MOVE THIS SOMEWHERE ELSE?}

@@

\section{Algorithm}

@@

    \begin{appendices}
        \section{Quantization}
        \FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.} \\
        \FIXME{PGR-TK Paper - Layouts of challenging regions of the human pangenome. - https://www.nature.com/articles/s41592-023-01914-y}
    \end{appendices}
